---
title: "Supply Chain Analytics with Vectors, Matrices, and Systems of Equations"
subtitle: "DATA 605 – Homework 1"
author: "Madina Kudanova"
format:
  html:
    embed-resources: true
    toc: true
execute:
  echo: true
---

```{r setup}
#| message: false
library(MASS)  # for ginv() – used in Part C
```

## Reading Data files (inventory.csv, demand.csv, and shipments.csv)

```{r data}
inventory <- read.csv("inventory.csv")
demand_df <- read.csv("demand.csv")
shipments <- read.csv("shipments.csv")

# convert demand.csv (long format: Region, Demand) into a named vector
demand <- setNames(demand_df$Demand, demand_df$Region)

inventory
demand
shipments

```

## Part A: Inventory Analysis (Vector Operations)

### 1. Total inventory per region

```{r part-a-totals}
region_totals <- c(
  Region1 = sum(inventory$Region1),
  Region2 = sum(inventory$Region2),
  Region3 = sum(inventory$Region3)
)
region_totals
```

### 2. Surplus or shortage relative to demand

```{r part-a-surplus}
surplus <- region_totals - demand
surplus
```

### 3.Region most at risk

```{r part-a-risk}
at_mostrisk <- names(surplus)[which.min(surplus)]
at_mostrisk
```

```{r part-a-plot}
#| fig-cap: "Inventory vs. Demand by Region"
bar_data <- rbind(Inventory = region_totals, Demand = demand)
barplot(bar_data, beside = TRUE, col = c("blue", "firebrick"),
              legend.text = TRUE, main = "Inventory vs. Demand by Region",
        ylab = "    Units")
```

### 4. Interpretaton of business implication

Region1 and Region2 each have 300 more units than needed. Region3 has exactly 2,500 units of inventory against 2,500 in demand, meaning no extra units are available. This makes Region3 the most at risk of running short if demand increases or a shipment is delayed. Management should prioritize replenishing Region3, reevaluate its demand forecast, and consider moving some surplus inventory from Region1 or Region2 to Region3 instead of ordering new stock.

## Part B: Shipment Matrix Analysis (Matrix Operations)

### 1. Shipment matrix

```{r part-b-matrix}
Shipment_matrix <- as.matrix(shipments[, c("Region1", "Region2", "Region3")])
rownames(Shipment_matrix) <- shipments$Warehouse
Shipment_matrix
```

### 2. Row totals (shipped by warehouse) and column totals (received by region)

```{r part-b-totals}
row_totals <- rowSums(Shipment_matrix)

col_totals <- colSums(Shipment_matrix)
row_totals
col_totals
```

### 3. Warehouse shipping the most &
### 4. Region receiving the most

```{r part-b-max}
top_warehouse <- names(row_totals)[which.max(row_totals)]
top_region    <- names(col_totals)[which.max(col_totals)]
top_warehouse
top_region
```

```{r part-b-plot}
#| fig-cap: "Shipments by Warehouse and Region"
barplot(Shipment_matrix, beside = TRUE, col = c("magenta", "darkorange", "green"),
        legend.text = rownames(Shipment_matrix), main = "Shipments by Warehouse and Region",
        ylab = "Units Shipped")
```

## Part C: Resource Allocation Model 

Management wants to determine how many additional truckloads should be
sent from each warehouse next week to meet projected demand. Let:

- $x$ = truckloads shipped from Warehouse A
- $y$ = truckloads shipped from Warehouse B
- $z$ = truckloads shipped from Warehouse C

Based on transportation capacity and regional replenishment requirements,
management developed the following system of equations:

$$
\begin{aligned}
x + y + z &= 18 \\
2x + y &= 26 \\
y + 2z &= 22
\end{aligned}
$$

These equations represent:

- Total truckloads available for shipment
- Capacity constraints at Warehouse A
- Combined delivery requirements for Warehouses B and C

### Task 1: Express the System in Matrix Form

Rewriting the system as $Ax = b$:

$$
\begin{bmatrix} 1 & 1 & 1 \\ 2 & 1 & 0 \\ 0 & 1 & 2 \end{bmatrix}
\begin{bmatrix} x \\ y \\ z \end{bmatrix}
=
\begin{bmatrix} 18 \\ 26 \\ 22 \end{bmatrix}
$$

Where:

- **Coefficient matrix** $A$ (3×3, from the coefficients of x, y, z in each equation)
- **Variable vector** $x = \begin{bmatrix} x \\ y \\ z \end{bmatrix}$ (truckloads from Warehouse A, B, C)
- **Constants vector** $b = \begin{bmatrix} 18 \\ 26 \\ 22 \end{bmatrix}$

```{r part-c-setup}
A <- matrix(c(1, 1, 1,
              2, 1, 0,
              0, 1, 2),
            nrow = 3, byrow = TRUE)
b <- c(18, 26, 22)
A
b
```

### Task 2: Solve the System Computationally

Before solving, we need to check whether an exact solution exists:

```{r part-c-check}
det(A)
rankA   <- qr(A)$rank
rankAug <- qr(cbind(A, b))$rank
rankA
rankAug
```

`det(A) = 0` means $A$ is singular. `rank(A) = 2` while `rank([A|b]) = 3`
means the system is inconsistent — no exact (x, y, z) can satisfy all
three equations. Attempting a direct solve confirms this:

```{r part-c-solve}
#| error: true
solve(A, b)
```

**Algebraic proof:** take $2 \times$ Equation 1 minus Equation 2:

$$
2(x+y+z) - (2x+y) = 2(18) - 26 \;\Rightarrow\; y + 2z = 10
$$

But Equation 3 independently states $y + 2z = 22$ — a direct
contradiction, so no (x, y, z) satisfies all three equations
simultaneously.

The closest achievable compromise (least-squares) is computed instead:

```{r part-c-lsq}
solution <- as.vector(ginv(A) %*% b)
names(solution) <- c("x", "y", "z")
round(solution, 2)
```

**Values obtained:** $x \approx 8.33$, $y \approx 7.33$, $z \approx 6.33$.

### Task 3: Verify the Solution

```{r part-c-verify}
check <- A %*% solution
round(check, 2)

residual <- as.vector(check) - b
round(residual, 2)
```

The residuals (+4, -2, -2) are not zero, so $Ax \neq b$ — confirming the
least-squares values are only the closest fit, not an exact solution.

```{r part-c-plot}
#| fig-cap: "Least-Squares Truckload Allocation by Warehouse"
barplot(solution, col = c("cyan", "chartreuse", "coral"),
        main = "Recommended Additional Truckloads by Warehouse (least-squares)",
        ylab = "Truckloads")
```

### Task 4: Business Interpretation

Region1 and Region2 each have 300 more units than needed. Region3 has
exactly 2,500 units of inventory against 2,500 in demand, meaning no
extra units are available (Part A). Warehouse A ships the most product
overall (1,200 units), and Region1 receives the most (1,200 units)
(Part B). This system has no exact solution: no combination of
truckloads from A, B, and C satisfies all three planning targets at
once. The closest achievable values are approximately 8, 7, and 6
truckloads from Warehouses A, B, and C respectively, with Warehouse A
still contributing the most.

Because no exact allocation exists, management cannot simply solve its
way to a plan; at least one of the three targets has to be relaxed. The
gap is largest around Warehouses B and C's combined requirement, which
lines up with Region3 (served heavily by those two warehouses) already
sitting at zero surplus in Part A.

This is why systems of linear equations are useful in supply chain
planning: they test whether a set of operational targets is achievable
together at all, before trucks are dispatched.